---
title: "PBMC 3k: QC and clustering"
format:
  html:
    toc: true
    embed-resources: true
---

## What this is

A short write-up of the PBMC 3k walkthrough. It reads the results produced by
`scripts/01_qc_cluster.R` and explains each step. In the full system this same source renders twice,
once as a learning version with the code and concept notes shown, and once as a clean report with the
code folded and a provenance footer. Here it is kept as one simple document.

Every number below is read from a file in `results/`. None of it is typed in by hand.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
qc <- read.csv("results/qc_summary.csv")
markers <- read.csv("results/marker_avg_expression.csv", row.names = 1)
```

## Quality control

Cells were kept using the thresholds pre-registered at the design step: 200 to 2500 genes, and
mitochondrial content below 5 percent. The filter logic is a unit-tested pure function rather than
inline code, so the decision that sets the numbers is verified on its own.

```{r qc}
qc
```

`r qc$n_cells_pass` of `r qc$n_cells_loaded` cells passed, and clustering at resolution 0.5 gave
`r qc$n_clusters` clusters.

## Does it make biological sense?

The check is whether the canonical PBMC markers land where the biology says they should.

```{r markers}
knitr::kable(round(markers, 2))
```

High CD3D marks the T cells, MS4A1 the B cells, CD14 the monocytes, NKG7 the NK cells, and PPBP the
platelets. If those line up, the clustering is believable. If they scatter, the loop goes back a step
before anything is trusted.

## Reproducibility

Re-running `scripts/01_qc_cluster.R` with the same seed gives the same result. In a real project this
runs inside a pinned container, and the report carries a provenance footer with the commit and the
environment.
